Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-06-09 paper.Rnw
put eq. 9 into eq. 8 #7

Below, we describe a Bayesian PS model fit in one step.
However, the full process of PS analysis is a much more involved
procedure.
After a formal introduction in Section \ref{sec:psIntro},
\S~\ref{sec:psModel} describes a set of sub-models for student hint
requests, posttest scores, and treatment effects.
Ideally, each of these models should be carefully tailored to each
data application, and each must be rigorously tested.
Model checking includes examination of residual data plots, fake data
simulation, and fitting a range of alternative specifications in a
sensitivity analysis.
More details for checking PS models can be found in \citet{aoas},
\citet{psTutorial}, and for Bayesian models in general in
\citet{gelman2014bayesian} and \citet{mcelreath2020statistical}.

@@

For those reasons, we modeled hint usage at the problem level, with
equation \eqref{eq:rasch}, but with two important differences.
First, a conceptual difference: the $\eta$ in \eqref{eq:rasch} was
replaced by $\eta (1)$, indicating that it is measuring
\emph{potential} hint usage---the hint usage a student would exhibit
were he or she assigned to the treatment condition.
Whereas $\eta$ is only defined for students assigned to the treatment condition,
$\eta(1)$ is defined for all the students in the study.
Second, a modeling difference: since we did not dichotomize $\eta(1)$
in this analysis, we modeled it as normal , not as a normal mixture, as in
\eqref{eq:mixture}.
With this change, the PS estimand became
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
the treatment effect for students whose potential hint usage, if assigned to the treatment condition $\eta(1)=r$.
A full treatment of this variant of PS, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

We modeled students' posttest scores $Y$ as
conditionally normal:
\begin{equation}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
\gamma_{0b[i]}+\bm{x}_i^T\bm{\gamma}+a\eta(1)+Z_i\tau(\eta(1))+\zeta^{tch}_{t[i]}+\zeta^{scl}_{s[i]},\omega_{Z[i]}\right)
\end{equation}
where $\gamma_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\gamma}$ are the
covariate coefficients, and $\zeta^{tch}$, and
$\zeta^{scl}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\omega$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

Finally, we modeled treatment effects
$\tau(\eta(1))$ as linear:
\begin{equation}\label{eq:tauModel}
\EE[Y(1)-Y(0)|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)
\end{equation}
While more complex models for $\tau(\eta(1))$ are theoretically
possible (for instance, \citet{jin2008principal} uses a quadratic
model), in our experience non-linear effect models do not perform as
well in model checks as the the linear model (\ref{eq:tauModel}).

>>>>

Below, we describe a Bayesian PS model fit in one step.
However, the full process of PS analysis is a much more involved
procedure.
After a formal introduction in Subsection \ref{sec:psIntro},
Subsection \ref{sec:psModel} describes a set of sub-models for student hint
requests, posttest scores, and treatment effects.
Ideally, each of these models should be carefully tailored to each
data application, and each must be rigorously tested.
Model checking includes examination of residual data plots, fake data
simulation, and fitting a range of alternative specifications in a
sensitivity analysis.
More details for checking PS models can be found in \citet{aoas},
\citet{psTutorial}, and for Bayesian models in general in
\citet{gelman2014bayesian} and \citet{mcelreath2020statistical}.

@@

For those reasons, we modeled hint usage at the problem level, with
equation \eqref{eq:rasch}, but with two important differences.
First, a conceptual difference: the $\eta$ in \eqref{eq:rasch} was
replaced by $\eta (1)$, indicating that it is measuring
\emph{potential} hint usage---the hint usage a student would exhibit
were he or she assigned to the treatment condition.
Whereas $\eta$ is only defined for students assigned to the treatment condition,
$\eta(1)$ is defined for all the students in the study.
Second, a modeling difference: instead of the normal mixture model
\eqref{eq:mixture}, we modeled $\eta(1)$ as normal.
With this change, the PS estimand became
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
the treatment effect for students whose potential hint usage, if assigned to the treatment condition $\eta(1)=r$.
A full treatment of this variant of PS, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

We modeled students' posttest scores $Y$ as
conditionally normal:
\begin{equation}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
\gamma_{0b[i]}+\bm{x}_i^T\bm{\gamma}+a\eta(1)+Z_i(b_0+b_1\eta(1))+\zeta^{tch}_{t[i]}+\zeta^{scl}_{s[i]},\omega_{Z[i]}\right)
\end{equation}
where $\gamma_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\gamma}$ are the
covariate coefficients, and $\zeta^{tch}$, and
$\zeta^{scl}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\omega$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

Model \eqref{eq:outcomeSubmodel} implies that treatment effects are
linear in $\eta(1)$,
$\EE[Y(1)-Y(0)|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)$

While more complex models for $\tau(\eta(1))$ are theoretically
possible (for instance, \citet{jin2008principal} uses a quadratic
model), in our experience non-linear effect models do not perform as
well in model checks as the the linear model.
